Alternating Automata Modulo First Order Theories

### 2019-05-21 quantifiers/cav2019/body.tex
added final version

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word from $\lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches. Termination of lazy annotation
procedures is not guaranteed, but having a suitable coverage relation
between the nodes of the search tree may ensure convergence of many
real-life examples. However, applying lazy annotation to first order
alternating automata faces two nontrivial
problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, in
  general, to decide if a path is infeasible. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

\paragraph{\em Applications}
We use first order alternating automata to develop practical
semi-algorithms for a number of known undecidable problems, such as:
inclusion of regular timed languages \cite{AlurDill94}, inclusion of
quasi-regular languages recognized by finite-memory automata
\cite{KaminskiFrancez94} and emptiness of predicate automata, a
subclass of first-order alternating automata used to verify
parameterized concurrent programs \cite{Farzan15,Farzan16}.

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event, if it occurs at all, it occurs after at
least $2$ time units. On the other hand, $B$ recognizes the sequences
in which there exists an event $a$ such that no event occurs after it
in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_2-t_1 \geq
  2}$ and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1},
  a, t_i, \sigma_{i+1}, t_{i+1}, \ldots \sigma_n, t_n \mid \sigma_1,
  \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in
  \real,~ 1 \leq i < n \Rightarrow t_{i+1}-t_i \neq 1}$. It is
proved\footnote{See \cite[Theorem 1]{AlurMadhusudan04} for a proof.}
that $\clang{B}$ cannot be recognized by a timed automaton, thus the
inclusion cannot be proved in the framework of timed automata
\cite{AlurDill94,AlurMadhusudan04}, by showing that $\lang{A} \cap
\clang{B} = \emptyset$.

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the (last) time value
time when the $\mathsf{x}$ clock of $A$ was reset (if it is reset
more than once) and $z$ is the time value of the last event. The time
stamp of the current event $a$ is the input variable $t$, which must
be bigger or equal to $z$, since the lapse of time between two
consequtive events is always positive. The current value of the
$\mathsf{x}$ clock is thus $z-x$ and the guard $\mathsf{x} \geq 2$ on
the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes $z-x \geq 2$
in the second rule. The reset of $x$ on the $s_1 \arrow{}{} s_2$
transition is encoded by passing $z$ as the first parameter of $s_2$,
in the first rule. The initial configuration of $A$ [$B$] is
$s_1(0,0)$ [$q_1(0,0)$] and $s_2$ [$q_2$] marks the final
configurations.

\begin{figure}[htb]
\vspace*{-2\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
with final configurations marked by $q_1$, instead of $q_2$. Finally,
we check the emptiness of the alternating automaton consisting of the
rules of $A$ and $\overline{B}$, started in the initial configuration
$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
emptiness of this automaton in $\sim\!0.3$ seconds, on an average
laptop machine.

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
extends the alternating automata with variables ranging over infinite
data considered in \cite{IosifXu18}. There all variables were required
to be observable in the input. We overcome this restriction by
allowing internal (invisible) variables. Another closely related work
\cite{IosifRV16} considers an inclusion between an asynchronous
product of automata $A_1 \times \ldots \times A_n$, extended with data
variables, and a monitor automaton $B$. The semi-algorithm defined
there was based on the assumption that all variables of the observer
$B$ must be declared in the automata $A_1, \ldots, A_n$ under
check. First order alternating automata bypass this limitation, since
our emptiness checking semi-algorithm can handle invisible variables.

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Since thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or use quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

For space reasons, the proofs of the technical results are given in
Appendinx \ref{app:proofs}.

\vspace*{-.5\baselineskip}
\subsection{Preliminaries}

@@

We consider the interpretation of all function symbols $f :
\Data^{\#(f)} \rightarrow \Data$ to be fixed by the interpretation of
the $\Data$ sort, for instance if $\Data$ is the set of integers
$\zed$, these are zero, the successor function and the arithmetic
operations of addition and multiplication. We extend this convention
to several predicates over $\Data$, such as the inequality relation
over $\zed$, and write $\preds$ for the set of \emph{uninterpreted
  predicates}.

Let $\vars = \set{x,y,z,\ldots}$ be a countably infinite set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
formulae is defined by the syntax below:
\[\phi := t = s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms and $p$ is a
predicate symbol. We write $\phi_1 \vee \phi_2$, $\phi_1 \rightarrow
\phi_2$ and $\forall x ~.~ \phi_1$ for $\neg(\neg\phi_1 \wedge
\neg\phi_2)$, $\neg\phi_1 \vee \phi_2$ and $\neg\exists x ~.~
\neg\phi_1$, respectively. $\fv{}{\phi}$ is the set of free variables
in $\phi$ and the size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down. A \emph{sentence} is a formula
$\phi$ with no free variables. A formula is \emph{positive} if each
uninterpreted predicate symbol occurs under an even number of
negations and we denote by $\posforms(Q,X)$ the set of positive
formulae with predicates from the set $Q \subseteq \preds$ and free
variables from the set $X \subseteq \vars$. A formula is in
\emph{prenex form} if it is of the form $\varphi = Q_1x_1 \ldots
Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In this case we
call $\phi$ the \emph{matrix} of $\varphi$. Every first order formula
can be written in prenex form, by renaming each quantified variable to
a unique name and moving the quantifiers upfront.

@@

An \emph{interpretation} $\I$ maps each predicate symbol $p$ into a
set $p^\I \subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element
of $\Bool$ if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable
$x$ into an element of $\Data$. Given a term $t$, we denote by $t^\nu$
the value obtained by replacing each variable $x$ by the value
$\nu(x)$ and evaluating each function application. For a formula
$\phi$, we define the forcing relation $\I,\nu \models \phi$
recursively on the structure of $\phi$, as usual.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ifLongVersion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere.  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
for sentences. A sentence $\phi$ is \emph{satisfiable} if
$\sem{\phi}\neq\emptyset$. An element of $\sem{\phi}$ is called a
\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
\models \phi$ for every interpretation $\I$ and every valuation
$\nu$. We say that $\phi$ \emph{entails} $\psi$, written $\phi \models
\psi$ if and only if $\sem{\phi} \subseteq \sem{\psi}$.

@@

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules}. A transition rule is of
the form \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in
Q$ is a predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
of generality, we consider, for each predicate $q \in Q$ and each
input event $a \in \Sigma$, at most one such rule, as two or more
rules can be joined using disjunction. The quantifiers occurring in
the right-hand side formula of a transition rule are called
\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

The semantics of first order alternating automata is analogous to the
semantics of propositional alternating automata, with rules of the
form $q \arrow{a}{} \phi$, where $q$ is a propositional variable and
$\phi$ a positive boolean combination of propositional variables. For
instance, $q_0 \arrow{a}{} (q_1 \wedge q_2) \vee q_3$ means that the
automaton can choose to transition in either both $q_1$ and $q_2$ or
in $q_3$ alone. This leads to defining transitions as the
\emph{minimal models} of the right hand side of a rule\footnote{Both
  $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
    \bot}$ and $\set{q_1 \leftarrow \bot, q_2 \leftarrow \bot, q_3
    \leftarrow \top}$ are minimal models, however $\set{q_1 \leftarrow
    \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$ is a model but is
  not minimal.}. The original definition of alternating automata
\cite{ChandraKozenStockmeyer81} works around this problem and
considers boolean valuations instead of formulae. In contrast, a
finite description of a first order alternating automaton cannot be
given in terms of interpretations, as a first order formula may have
infinitely many models, corresponding to infinitely many initial or
successor states occurring within an execution step.

@@

An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$.

\begin{example}\label{ex:foada}
  Consider the automaton $\A = \tuple{\set{a}, \set{x},
    \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
  is the set: \(q_0(y) \arrow{a(x)}{} q_1(y+x) \wedge q_2(y-x)\),
  \(q_1(y) \arrow{a(x)}{} q_1(y+x) \vee (y > 0 \wedge q_f)\) and
  \(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A
  possible execution tree of this automaton is the following:
  \begin{center}
    \input exec.pdf_t
  \end{center}
  The execution tree is not accepting, since its frontier is not
  labeled with final configurations everywhere. Incidentally, here we
  have $\lang{\A} = \emptyset$, which is proved by our tool in
  $\sim\!0.5$ seconds on an average machine. \hfill$\blacksquare$
\end{example}

In the rest of this paper, we are concerned with the following
problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under complement and,
further, set up the ground for developping a practical semi-algorithm
for the emptiness problem.

@@

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
  subformula of $\pathform{\event{w}}$ corresponding to the
  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
  interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

>>>>

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word from $\lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches. Termination of lazy annotation
procedures is not guaranteed, but having a suitable coverage relation
between the nodes of the search tree may ensure convergence of many
real-life examples. However, applying lazy annotation to first-order
alternating automata faces two nontrivial
problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, in
  general, to decide if a path is infeasible. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first-order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

We use first-order alternating automata to develop practical
semi-algorithms for a number of known undecidable problems, such as:
inclusion of regular timed languages \cite{AlurDill94}, inclusion of
quasi-regular languages recognized by finite-memory automata
\cite{KaminskiFrancez94} and emptiness of predicate automata, a
subclass of first-order alternating automata used to verify
parameterized concurrent programs \cite{Farzan15,Farzan16}.

\paragraph{\em Related Work}
Recognizers for languages over infinite alphabets have found various
applications, ranging from Unicode text recognition
\cite{DAntoniKincaidWang18} to runtime program monitoring
\cite{BarringerRydeheardHavelund07}. Extending finite automata to
infinite alphabets has been considered in the context of
\emph{symbolic alternating finite automata} (s-AFA), whose transitions
are labeled with guards taken from a decidable theory of the data
domain \cite{DAntoniKincaidWang18}. As in our model, s-AFA are closed
under union, intersection and complement and emptiness is decidable,
due to the lack of registers. However, s-AFA are strictly less
expressive than our model, because comparing data at different
positions in the input word is not possible.

\emph{Constrained Horn clauses} (CHC) are a branching computation
model widespread in program verification \cite{Grebenshchikov12}. The
main difference between alternating and bottom-up branching
computations is that, in an alternating model, all branches of the
computation must synchronize on the same input word. With this in
mind, it is possible to express emptiness of first-order alternating
automata as the existence of solutions of a CHC over a higher-order
theory of data, extended with algebraic data types (lists). The
effectiveness of such an encoding depends on the effectiveness of
interpolation and witness term generation for theories of algebraic
data types \cite{HojjatRuemmer18}.

The alternating automata model presented in this paper extends the
alternating automata with variables ranging over infinite data
considered in \cite{IosifXu18}. There all variables were required to
be observable in the input. We overcome this restriction by allowing
internal (invisible) variables. Another closely related work
\cite{IosifRV16} considers an inclusion between an asynchronous
product of automata $A_1 \times \ldots \times A_n$, extended with data
variables, and a monitor automaton $B$. The semi-algorithm defined
there was based on the assumption that all variables of the observer
$B$ must be declared in the automata $A_1, \ldots, A_n$ under
check. This limitation can now be bypassed, since the inclusion
problem can be encoded as emptiness of a first-order alternating
automaton and, moreover, the emptiness checking semi-algorithm can
handle invisible variables.

The work probably closest to ours concerns the model of
\emph{predicate automata} (PA) \cite{Farzan15,Farzan16,KincaidPhD},
used in the verification of parameterized concurrent programs with
shared memory. In this model, the alphabet consists of pairs of
program statements and thread identifiers and is considered infinite
because the number of threads is unbounded. Since thread identifiers
can only be compared for equality, the data theory in PA is the theory
of equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or use quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permutations of
argument values. However, no semi-algorithm has been given for
quantified PA. Dealing with quantified transition rules is one of our
contributions.

\subsection{Preliminaries}

@@

We consider the interpretation of all function symbols $f :
\Data^{\#(f)} \rightarrow \Data$ to be fixed by the interpretation of
the $\Data$ sort, for instance if $\Data$ is the set of integers
$\zed$, these are zero, the successor function and the arithmetic
operations of addition and multiplication. We extend this convention
to several predicates over $\Data$, such as the inequality relation
over $\zed$, and write $\preds$ for the set of remaining
\emph{uninterpreted predicates}.

Let $\vars = \set{x,y,z,\ldots}$ be a countably infinite set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t = s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms and $p$ is a
predicate symbol. We write $\phi_1 \vee \phi_2$, $\phi_1 \rightarrow
\phi_2$ and $\forall x ~.~ \phi_1$ for $\neg(\neg\phi_1 \wedge
\neg\phi_2)$, $\neg\phi_1 \vee \phi_2$ and $\neg\exists x ~.~
\neg\phi_1$, respectively. $\fv{}{\phi}$ is the set of free variables
in $\phi$ and the size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down. A \emph{sentence} is a formula
$\phi$ with no free variables. A formula is \emph{positive} if each
uninterpreted predicate symbol occurs under an even number of
negations and we denote by $\posforms(Q,X)$ the set of positive
formulae with predicates from the set $Q \subseteq \preds$ and free
variables from the set $X \subseteq \vars$. A formula is in
\emph{prenex form} if it is of the form $\varphi = Q_1x_1 \ldots
Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In this case we
call $\phi$ the \emph{matrix} of $\varphi$. Every first-order formula
can be written in prenex form, by renaming each quantified variable to
a unique name and moving the quantifiers upfront.

@@

An \emph{interpretation} $\I$ maps each predicate symbol $p$ into a
set $p^\I \subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element
of $\Bool$ if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable
$x$ into an element of $\Data$. Given a term $t$, we denote by $t^\nu$
the value obtained by replacing each variable $x$ by the value
$\nu(x)$ and evaluating each function application. For a formula
$\phi$, we define the forcing relation $\I,\nu \models \phi$
recursively on the structure of $\phi$, as usual. For a formula
$\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu \isdef \set{\I
  \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript for
sentences. A sentence $\phi$ is \emph{satisfiable} if
$\sem{\phi}\neq\emptyset$. An element of $\sem{\phi}$ is called a
\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
\models \phi$ for every interpretation $\I$ and every valuation
$\nu$. We say that $\phi$ \emph{entails} $\psi$, written $\phi \models
\psi$ if and only if $\sem{\phi} \subseteq \sem{\psi}$.

@@

A \emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states and
$\Delta$ is a set of \emph{transition rules}. A transition rule is of
the form \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in
Q$ is a predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
of generality, we consider, for each predicate $q \in Q$ and each
input event $a \in \Sigma$, at most one such rule, as two or more
rules can be joined using disjunction. The quantifiers occurring in
the right-hand side formula of a transition rule are called
\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

The semantics of first-order alternating automata is analogous to the
semantics of propositional alternating automata, with rules of the
form $q \arrow{a}{} \phi$, where $q$ is a propositional variable and
$\phi$ a positive boolean combination of propositional variables. For
instance, $q_0 \arrow{a}{} (q_1 \wedge q_2) \vee q_3$ means that the
automaton can choose to transition in either both $q_1$ and $q_2$ or
in $q_3$ alone. This leads to defining transitions as the
\emph{minimal models} of the right hand side of a rule\footnote{Both
  $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
    \bot}$ and $\set{q_1 \leftarrow \bot, q_2 \leftarrow \bot, q_3
    \leftarrow \top}$ are minimal models, however $\set{q_1 \leftarrow
    \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$ is a model but is
  not minimal.}. The original definition of alternating automata
\cite{ChandraKozenStockmeyer81} works around this problem and
considers boolean valuations instead of formulae. In contrast, a
finite description of a first-order alternating automaton cannot be
given in terms of interpretations, as a first-order formula may have
infinitely many models, corresponding to infinitely many initial or
successor states occurring within an execution step.

@@

An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
and only if all paths in $\T$ have the same length and the frontier of
each tree $T \in \T$ is labeled with final configurations.  If $\A$
has an accepting execution over $w$ starting with a cube $c \in
\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
$\lang{\A}$ be the set of words accepted by $\A$. For example,
consider the automaton $\A = \tuple{\set{a}, \set{x},
  \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
is the set: \(q_0(y) \arrow{a(x)}{} q_1(y+x) \wedge q_2(y-x)\),
\(q_1(y) \arrow{a(x)}{} q_1(y+x) \vee (y > 0 \wedge q_f)\) and
\(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A possible
execution tree of this automaton is the following:
\begin{center}
  \input exec.pdf_t
\end{center}
The execution tree is not accepting, since its frontier is not
labeled with final configurations everywhere. Incidentally, here we
have $\lang{\A} = \emptyset$, which is proved by our tool in
$\sim\!0.5$ seconds on an average machine.

\noindent
In the rest of this paper, we are concerned with the following
problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first-order alternating automata are closed under complement and,
further, set up the ground for developping a practical semi-algorithm
for the emptiness problem.

@@

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
